\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel-space

### 2012-05-18 design.tex
copyedit

In this work, we focus specifically on nondeterministic scheduling driven by timer interrupts. Landslide assumes that timer interrupts are the only source of nondeterminism for the guest kernel, so controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case. In future work (Section~\ref{sec:future-drivers}), we may also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.

@@

\begin{itemize}
	\item {\bf Iterative refinement of interleaving granularity.} The recommended usage pattern is to begin by searching state spaces with coarse-grained interleavings, and then refining the decision set afterwards until bugs are found. Landslide automatically identifies a ``minimal decision set'', comprising only voluntary reschedules (Section~\ref{sec:components-arbiter}), to use as the coarsest granularity.
	\item {\bf Decision point selection.} Landslide provides an interface for the user to manually limit identification of decision points to only ``relevant'' modules of the kernel, thereby only generating thread interleavings around points within those modules.
	\item {\bf Memory conflict selection.} Landslide also provides a mechanism for ignoring shared memory conflicts on certain common data structures (especially those accessed in every thread transition, such as the scheduler queues), to enable DPOR to achieve greater reduction. Intuitively, this represents sacrificing the ability to find races around those data structures to make searching other components more feasible.
\end{itemize}

We advocate relying on the user for such information because the user need only understand the basics of their kernel design (which is difficult for Landslide to guess) while not knowing in advance what bugs are being looked for (which Landslide is able to provide). We claim that this combination of the user's knowledge and Landslide's testing mechanisms leads to an effective usage dynamic, in which the user ``steers'' Landslide towards focused search spaces that are more likely to find bugs.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.65\textwidth]{landslide.png}
	\caption{A simple visualisation of Landslide and its components, and how they interact with the guest kernel.}
	\label{fig:landslide}
\end{figure}

@@

The Landslide scheduler also tracks which important actions each thread is performing. These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section \revision{(note that ``vanish'' is simply Pebbles's name for ``exit'')}, and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing (voluntary reschedule, etc), and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
There are also other minor actions: {\em readlining}, used to track test lifecycle (Section~\ref{sec:components-test}), {\em just\_forked}, used for special-case context switch behaviour (Section~\ref{sec:components-arbiter}), and also several flags for tracking mutex lock/unlock events.}

@@

Landslide maintains a mirror image of the guest kernel's dynamic allocation heap, so it can know at any point which memory ranges are allocated and which ranges used to be allocated but now are freed. This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.
\revision{This heap tracking provides the ability to check for dynamic allocation errors (e.g., use-after-free and double-free bugs), such as in Valgrind~\cite{valgrind}.}

@@

In Landslide, it is always necessary to identify a decision point during a voluntary reschedule, to maintain the invariant that each transition between decision points is comprised of the execution of only one thread.
\revision{In future work, we may also investigate the implications of allowing some transitions to consist of multiple threads' executions, which would have applications for heuristic state space reduction.}

\revision{
Because voluntary reschedules happen of the guest kernel's own volition, when a thread begins one, it will necessarily not have been preempted by the timer handler. Hence, we identify voluntary reschedules when a thread enters the context switcher without having entered the timer handler.\footnote{
Landslide actually identifies voluntary reschedules at the point when the reschedule ends, so the {\em previous} thread would not have entered the timer handler, and the current thread may or may not have.
We do this to avoid the complexity of dealing with the possibility that the guest kernel would have interrupts disabled at the beginning of the reschedule and not re-enable them until after the thread switch.}
}

@@

\begin{itemize}
	\item {\bf Is the shell waiting for keyboard input?} In general, the shell only waits for keyboard input when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process might not yet have reaped a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well. For these cases an additional check is needed.
	\item {\bf Is the kernel idling?} This indicates whether any work is left to be done during bootup or during teardown.
	\item {\bf How many threads currently exist on the system?} This indicates whether the test has begun running.
\end{itemize}

The states in the test lifecycle, and the rules for transitioning between them, are as follows.

@@

Figure~\ref{fig:teststate} depicts the states and transitions visually.

@@

\revision{Instead, Landslide uses considers every evil ancestor of each transition when performing DPOR. This modification enables Landslide to avoid the problem presented in the figure, because it will identify both alternate interleavings after exploring the first one. We claim without certainty that this change restores the second notion of soundness to Landslide's search, and leave proving this property to future work.}

@@

\begin{enumerate}
	\item If the kernel has panicked, Landslide prints the message used in the panic/assertion.
	\item If a use-after-free bug is found, Landslide prints information about the most-recently-freed chunk containing the accessed address: the stack trace and TID both for when it was allocated and freed.
	\item If deadlock is detected, Landslide prints the cycle of TIDs that are blocked on each other.\footnote{This currently only works when using the annotations for yielding mutexes, though it is not difficult to implement more generally.}
	\item If a memory leak is suspected, Landslide prints how many bytes bigger the heap is after the test ended than when the test began.\footnote{Future feature: Printing when-allocated stack traces for each suspicious heap chunk.}
	\item If an infinite loop is suspected, depending on which heuristic was triggered, Landslide prints either the number of instructions since the last decision point (and the previous average) or the current branch depth (and the previous average).
\end{enumerate}

>>>>

In this work, we focus specifically on nondeterministic scheduling driven by timer interrupts. Landslide assumes that timer interrupts are the only source of nondeterminism for the guest kernel, so controlling when they occur theoretically allows for complete control over the concurrent behaviour of the test case. In future work (Section~\ref{sec:future-drivers}), we also address causes of non-determinism more complex than timer-driven thread scheduling, such as interrupts and data I/O from peripheral devices.

@@

\begin{itemize}
	\item {\bf Iterative refinement of interleaving granularity.} The recommended usage pattern is to begin by searching state spaces with coarse-grained interleavings, and then refining the decision set afterwards until bugs are found. Landslide automatically identifies a ``minimal decision set'', comprising only voluntary reschedules (Section~\ref{sec:components-arbiter}), to use as the coarsest granularity.
	\item {\bf Decision point selection.} Landslide provides an interface for the user to manually limit identification of decision points to only ``relevant'' modules of the kernel, thereby generating thread interleavings only around points within those modules.
	\item {\bf Memory conflict selection.} Landslide also provides a mechanism for ignoring shared memory conflicts on certain common data structures (especially those accessed in every thread transition, such as the scheduler queues), to enable DPOR to achieve greater reduction. Intuitively, this represents sacrificing the ability to find races around those data structures to make searching other components more feasible.
\end{itemize}

We advocate relying on the user for such information because the user need only understand the basics of their kernel design (which is difficult for Landslide to guess) while not knowing in advance which bugs are being looked for (which Landslide is able to provide). We claim that this combination of the user's knowledge and Landslide's testing mechanisms leads to an effective usage dynamic, in which the user ``steers'' Landslide towards focused search spaces that are more likely to find bugs.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{landslide.png}
	\caption{A simple visualisation of Landslide and its components, and how they interact with the guest kernel.}
	\label{fig:landslide}
\end{figure}

@@

The Landslide scheduler also tracks which important actions each thread is performing. These actions are {\em forking}, {\em vanishing}, and {\em sleeping}, which are described in the next section,\footnote{\revision{Note that ``vanish'' is simply Pebbles's name for ``exit''.}} and also {\em in\_timer\_handler} and {\em in\_context\_switch}, which express what type of context switch a given thread may be performing (voluntary reschedule, etc), and are useful for both thread scheduling (Section~\ref{sec:components-inflight}) and detecting voluntary reschedules (Section~\ref{sec:components-arbiter}).\footnote{
There are also other minor actions: {\em readlining}, used to track test lifecycle (Section~\ref{sec:components-test}), {\em just\_forked}, used for special-case context switch behaviour (Section~\ref{sec:components-arbiter}), and also several flags for tracking mutex lock/unlock events.}

@@

Landslide maintains a mirror image of the guest kernel's dynamic allocation heap, so it can know at any point which memory ranges are allocated and which ranges used to be allocated but now are freed. This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.
\revision{This heap tracking provides the ability to check for dynamic allocation errors (e.g., use-after-free and double-free bugs), in a similar fashion to Valgrind~\cite{valgrind}.}

@@

In Landslide, it is always necessary to identify a decision point during a voluntary reschedule, to maintain an invariant that each transition between decision points is comprised of the execution of only one thread.
\revision{In future work, we may also investigate the implications of allowing some transitions to consist of multiple threads' executions, which would have applications for heuristic state space reduction.}

\revision{
Because voluntary reschedules happen of the guest kernel's own volition, when a thread begins one, it will necessarily not have been preempted by the timer handler. Hence, we identify voluntary reschedules by determining when a thread enters the context switcher without having entered the timer handler.\footnote{
Landslide actually identifies voluntary reschedules at the point when the reschedule ends, so the {\em previous} thread would not have entered the timer handler, and the current thread may or may not have.
We do this to avoid the complexity of dealing with the possibility that the guest kernel would have interrupts disabled at the beginning of the reschedule and not re-enable them until after the thread switch.}
}

@@

\begin{itemize}
	\item {\bf Is the shell waiting for keyboard input?} In general, the shell waits for keyboard input only when the kernel is ready for the user (or Landslide) to cause a test to run, or after the test has finished and a new one could be run.
		However, the shell may sometimes be waiting for keyboard input while other threads have some work to do (for example, the init process might not yet have reaped a reparented child process), and this work may be affected by race conditions during the test, so we need to include it in the test as well. For these cases an additional check is needed.
	\item {\bf Is the kernel idling?} This indicates whether any work is left to be done during bootup or during teardown.
	\item {\bf How many threads currently exist on the system?} This indicates whether the test has begun running.
\end{itemize}

The states in the test lifecycle, and the rules for transitioning between them, are as follows.
Figure~\ref{fig:teststate} depicts the states and transitions visually.

@@

@@

\revision{Instead, Landslide considers every evil ancestor of each transition when performing DPOR. This modification enables Landslide to avoid the problem presented in the figure, because it will identify both alternate interleavings after exploring the first one. We claim without certainty that this change restores the second notion of soundness to Landslide's search, and leave proving this property to future work.}

@@

\begin{enumerate}
	\item If the kernel has panicked, Landslide prints the message used in the panic/assertion.
	\item If a use-after-free bug is found, Landslide prints information about the most-recently-freed chunk containing the accessed address: the stack trace and TID both for when it was allocated and freed.
	\item If deadlock is detected, Landslide prints the cycle of TIDs that are blocked on each other.\footnote{This currently works only when using the annotations for yielding mutexes, though it is not difficult to implement more generally.}
	\item If a memory leak is suspected, Landslide prints how many bytes bigger the heap is after the test ended than when the test began.\footnote{Future feature: Printing when-allocated stack traces for each suspicious heap chunk.}
	\item If an infinite loop is suspected, depending on which heuristic was triggered, Landslide prints either the number of instructions since the last decision point (and the previous average) or the current branch depth (and the previous average).
\end{enumerate}

### 2012-05-18 related.tex
remove some todos and fixmes
minor updates
copyedit

% FIXME: Give another pass for dad's feedback - another sentence about each thing saying what landslide does better

@@

Finally, DeMeter~\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique on only one component of the program at a time. Landslide makes use of the same general concept in its user interface (Chapter~\ref{sec:using}) by providing the user with options to constrain decision point identification to certain subsets of the kernel.

@@

SimTester~\cite{simtester} is another tool for testing kernel device drivers. Like Landslide, SimTester uses Simics as its execution environment, and identifies key decision points during execution at which to inject interrupts. SimTester focuses on races involving device interrupt handling code, whereas Landslide is geared towards non-driver inter-thread races. Additionally, SimTester's testing model involves forcing only one interrupt per test run, which is a lighter-weight approach than systematic exploration.

>>>>

@@

Finally, DeMeter~\cite{demeter} is a more recent tool for systematic exploration that introduces Dynamic Interface Reduction as a strategy for constraining the size of the state space, which means targetting the testing technique to only one component of the program at a time. Landslide makes use of the same general concept in its user interface (Chapter~\ref{sec:using}) by providing the user with options to constrain decision point identification to certain subsets of the kernel.

@@

\revision{
SimTester~\cite{simtester} is another tool for testing kernel device drivers. Like Landslide, SimTester uses Simics as its execution environment, and identifies key decision points during execution at which to inject interrupts. SimTester focuses on races involving device interrupt handling code, whereas Landslide is geared towards non-driver inter-thread races. Additionally, SimTester's testing model involves forcing only one interrupt per test run, which is a lighter-weight approach than systematic exploration.
}

### 2012-05-18 terminology.tex
copyedit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Terminology}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\revision{The following terms refer to parts of Landslide's analysis, regardless of the kernel under test's implementation.}

@@

		Figure~\ref{fig:threadfork} depicts a decision tree for a simple race with three possible interleavings.
	\item {\bf Branch}:
		One particular execution of the test case; a set of decision points with exactly one transition between each of them, characterising a single interleaving of threads.
	\item {\bf Interleaving}:
		A less precise / more abstract term for a branch, or for a subset of transitions which make up a branch.
	\item {\bf Bug}: An execution state which Landslide identifies as incorrect, which is indicative of illegal behaviour. Each branch of the tree may or may not have a bug in it. See Section~\ref{sec:techniques-bugs}.
	\item {\bf Backtracking}:
		Performed at the end of each branch of the tree, when the test case has finished running. Landslide identifies a decision point from the current branch at which the \revision{next interleaving to explore diverges from the current one}, reverts the state of the guest kernel to what it was at that decision point, and causes a different thread to run, hence exploring a different branch/interleaving.
	\item {\bf Decision trace}:
		A list of information about all decision points in a given branch, printed when a bug is found to help the user analyse its cause. Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace may also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
\end{enumerate}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Terminology}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{sec:terminology}

@@

\revision{The following terms refer to parts of Landslide's analysis, regardless of the implementation of the kernel under test.}

@@

		Figure~\ref{fig:threadfork} depicts a decision tree for a simple race with three possible interleavings.
	\item {\bf Branch}:
		One particular execution of the test case; a set of decision points with exactly one transition between each of them, characterising a single interleaving of threads.
	\item {\bf Interleaving}:
		A less precise / more abstract term for a branch, or for a subset of transitions which make up a branch.
	\item {\bf Bug}: An execution state which Landslide identifies as incorrect, which is indicative of illegal behaviour. Each branch of the tree may or may not have a bug in it. See Section~\ref{sec:techniques-bugs}.
	\item {\bf Backtracking}:
		Performed at the end of each branch of the tree, when the test case has finished running. Landslide identifies a decision point from the current branch at which the \revision{next interleaving to explore diverges from the current one}, reverts the state of the guest kernel to what it was at that decision point, and causes a different thread to run, hence exploring a different branch/interleaving.
	\item {\bf Decision trace}:
		A list of information about all decision points in a given branch, printed when a bug is found to help the user analyse its cause. Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace could also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
\end{enumerate}

### 2012-05-18 thesis.tex
fix some figs in eval
copyedit

\documentclass[12pt]{book}
%\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
\usepackage{upgreek}
%\usepackage{times}
\usepackage[charter]{mathdesign}
\usepackage{hyperref}
\usepackage{draftwatermark}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
%\usepackage{epsfig}
\usepackage[bottom]{footmisc}
\usepackage{mjz-titlepage}
\usepackage{framed}
\usepackage{setspace}
\setstretch{1.05}
\usepackage{subfig}
\usepackage{changebar}
\usepackage{colortbl}

@@

\revision{
Many techniques exist for dynamic testing of concurrent systems for race conditions.
Systematic exploration~\cite{verisoft}, the strategy we focus on in this work, involves making educated guesses at what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
Systematic exploration provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information. Compared to other dynamic analyses, such as data race detection~\cite{datacollider}, systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.

@@

We present Landslide,\footnote{
Landslide {\em (n)}: a phenomenon which demonstrates that Pebbles are not as stable as you might think.}
a tool for applying systematic race detection techniques in to kernel-level code.
It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
\revision{In order to use Landslide, a kernel programmer must instrument their kernel to inform Landslide of important concurrency events during the execution, and configure Landslide to focus its search on components of the kernel most relevant to the test.
%During a kernel's execution, Landslide records important actions performed by the kernel, decides at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
Using this instrumentation and configuration, Landslide is able to enumerate all possible interleavings of kernel threads at a granularity both fine-grained enough to expose interesting concurrency behaviour and coarse-grained enough to produce state spaces that are feasible to explore. Landslide then forces the kernel to exercise each of these interleavings, and checks for race conditions in each of them.}
When Landslide finds a bug (predicted by a set of common checks and heuristics), it stops execution and prints information about the sequence of interleavings that exposed the bug.

\revision{
In this work, we study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications (Chapter~\ref{sec:challenges}), present our techniques (some sound, and some heuristic) for addressing them (Chapter~\ref{sec:design}), and document the process of using Landslide (Chapter~\ref{sec:using}).
We evaluate Landslide both by studying its potential to be a helpful debugging aid in 15-410 and by studying certain bugs in detail to determine effective usage strategies (Chapter~\ref{sec:evaluation}).
Finally, we discuss possible avenues for future work in education, use of additional testing techniques, application to general purpose kernels, long-running testing approaches, and theoretical study of systematic exploration (Chapter~\ref{sec:future}).
}

>>>>

\documentclass[12pt]{book}
%\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
\usepackage{upgreek}
%\usepackage{times}
\usepackage[charter]{mathdesign}
\usepackage{hyperref}
\usepackage{draftwatermark}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
%\usepackage{epsfig}
\usepackage[bottom]{footmisc}
\usepackage{mjz-titlepage}
\usepackage{framed}
\usepackage{setspace}
\setstretch{1.05}
\usepackage{subfig}
\usepackage{changebar}
\usepackage{colortbl}
\usepackage{wrapfig}

@@

\revision{
Many techniques exist for dynamic testing of concurrent systems for race conditions.
Systematic exploration~\cite{verisoft}, the strategy we focus on in this work, involves making educated guesses as to what points during execution a preemption would be most likely to expose a bug, enumerating the different possibilities for interleaving threads around these points, and forcing the system to execute all such interleavings to check if any of them results in incorrect behaviour.
Systematic exploration provides a better alternative to conventional long-running stress tests, because it is less likely to overlook buggy execution patterns, and it enables a testing framework to report more thorough debugging information. Compared to other dynamic analyses, such as data race detection~\cite{datacollider}, systematic exploration is able to find a wider range of types of concurrency errors because of its ability to manipulate the execution of the system under test.

@@

We present Landslide,\footnote{
Landslide {\em (n)}: a phenomenon which demonstrates that Pebbles are not as stable as you might think.}
a tool for applying systematic race detection techniques to kernel-level code.
It is geared towards kernels that implement the Pebbles specification, the main project in Operating System Design and Implementation (15-410) at CMU, and is implemented as a module for Wind River Simics\textsuperscript{\texttrademark}~\cite{simics}, the x86 simulator that students use to run their Pebbles kernels.
\revision{In order to use Landslide, a kernel programmer must instrument their kernel to inform Landslide of important concurrency events during the execution, and configure Landslide to focus its search on components of the kernel most relevant to the test.
%During a kernel's execution, Landslide records important actions performed by the kernel, decides at which points in the kernel's execution a preemption will be most likely to expose a bug, and then exercises all possible interleavings of kernel threads around such points.
Using this instrumentation and configuration, Landslide is able to enumerate all possible interleavings of kernel threads at a granularity both fine-grained enough to expose interesting concurrency behaviour and coarse-grained enough to produce state spaces that are feasible to explore. Landslide then forces the kernel to exercise each of these interleavings, and checks for race conditions in each of them.}
When Landslide finds a bug (predicted by a set of common checks and heuristics), it stops execution and prints information about the sequence of interleavings that exposed the bug.

\revision{
In this work, we study the challenges inherent in applying systematic testing techniques in kernel-space, in contrast with user-space applications (Chapter~\ref{sec:challenges}), define common terms used in the rest of this document (Chapter~\ref{sec:terminology}), present our techniques (some sound, and some heuristic) for addressing them (Chapter~\ref{sec:design}), and document the process of using Landslide (Chapter~\ref{sec:using}).
We evaluate Landslide both by studying its potential to be a helpful debugging aid in 15-410 and by studying certain bugs in detail to determine effective usage strategies (Chapter~\ref{sec:evaluation}).
Finally, we discuss possible avenues for future work in education, use of additional testing techniques, application to general purpose kernels, long-running testing approaches, and theoretical study of systematic exploration (Chapter~\ref{sec:future}).
}
